---
title: "NYC Flights 2013"
author: "John Cruz"
date: "2023-02-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required Libraries
``` {r library, message=FALSE}
library(nycflights13)
library(tidyverse)
```

---

# Using dplyr

## Row Operations

### *filter()*

```{r filter}

# Flights that departed on January 1
flights |> 
  filter(month == 1 & day == 1)

# Flights that departed in January or February
flights |> 
  filter(month %in% c(1, 2))
```

### *arrange()*
``` {r arrange}
flights |> 
  arrange(desc(dep_delay))
```

### *distinct()*
``` {r distinct}
flights |> 
  distinct(origin, dest)
```

---

## Column Operations

### *mutate()*
- *.before* or *.after* "Determine new columns placement in data frame."
- *.keep* "Control which variables are kept. ('used' argument keeps the inputs from your calculations)"

``` {r mutate}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used"
  )
```

### *select()*
- starts_with("abc"): matches names that begin with “abc”.
- ends_with("xyz"): matches names that end with “xyz”.
- contains("ijk"): matches names that contain “ijk”.
- num_range("x", 1:3): matches x1, x2 and x3.

``` {r select}
# Select columns by name
flights |> 
  select(year, month, day)

# Select all columns between year and day (inclusive)
flights |> 
  select(year:day)

# Select all columns except those from year to day (inclusive)
flights |> 
  select(!year:day)

# Select all columns that are characters
flights |> 
  select(where(is.character))
```

### *rename()*
- alternative for many columns use *janitor::clean_names()*

``` {r rename}
flights |> 
  rename(tail_num = tailnum)
```

### *relocate()*

``` {r relocate}
flights |> 
  relocate(year:dep_time, .after = time_hour)
```

### *group_by()* & *summarize()*

``` {r groupby-summarize}
flights |> 
  group_by(month) |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE),
    count = n()
  )
```

### *summarize* based on tibble and multiple *group_by()*
When you are modifying a tibble that was created by multiple conditions in the previous group_by, you have to use .groups

``` {r summarize}
daily <- flights |>  
  group_by(year, month, day)

daily_flights <- daily |> 
  summarize(
    n = n(),
    .groups = "drop_last"
  )
```

### *slice_* functions
Instead of *n =*, you can use *prop = 0.1* to select 10% of the rows in each group.

- df |> slice_head(n = 1) takes the first row from each group.
- df |> slice_tail(n = 1) takes the last row in each group.
- df |> slice_min(x, n = 1) takes the row with the smallest value of x.
- df |> slice_max(x, n = 1) takes the row with the largest value of x.
- df |> slice_sample(n = 1) takes one random row.

``` {r slice-func}
flights |> 
  group_by(dest) |> 
  slice_max(arr_delay, prop = 0.05)
```






